---
title: "Maps"
output: 
  html_document: 
    keep_md: yes
---

In this tutorial we'll cover the following topics:

* Basic Maps
* ggplot2 Maps
* ggmap
* leaflet


# Basic Maps

There is no function for displaying maps in the base R functionality. To display
a simple map, we use the `maps` package. First, let's create a simple Indiana map.

```{r, warning=FALSE, message=FALSE}
library(maps)
map(database = 'state', regions = 'indiana')
```

Now we can add a few monitors as points. Use the longitude column
as the x-axis values and longitude for the y-axis values.

```{r}
monitors <- read.table(header=T, text='
  monitorid        lat        long   datum                    name            
          1   41.60668  -87.304729   WGS84              Gary-IITRI
          2  39.811097	-86.114469   WGS84  Indpls-Washington-Park
          3  39.749019	-86.186314   WGS84      Indpls-Harding-St.
          4  38.013248	-87.577856   WGS84  Evansville-Buena-Vista
          5  39.159383	-86.504762   WGS84             Bloomington
          6  39.997484  -86.395172   WGS84              Whitestown
                       ')
```


```{r, echo=FALSE}
map(database = 'state', regions = 'indiana')
points(x = monitors$long, y = monitors$lat)
```

We can jazz it up a bit by making it a county map and changing the symbol type and
color.

```{r}
map(database = 'county', regions = 'indiana')
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
```

We can also select specific counties to look at, and give the map a title.

```{r}
map(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
```

If we want the counties to be automatically labeled, we use the `map.text` 
function.

```{r}
map.text(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
```


And if we want to label the monitors, we use the `text()` function.

```{r}
map(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
text(x = monitors$long, y = monitors$lat, labels = monitors$name, pos = 2)
```



# ggplot2 Maps


The `ggplot2` package has a `geom_polygon()` function that by default draws a map
using lat/long data with a black fill. We first get the necessary data frame for
drawing a map of the U.S. from a `ggplot2` function named `map_data()`

```{r}
library(ggplot2)

us_data <- map_data("usa")
head(us_data)
```

```{r, message=FALSE, warning=FALSE}

ggplot() + geom_polygon(data = us_data, aes(x=long, y = lat, group = group))

```


We can recreate the Indiana county map with monitors by using the `"county"`
option in the `map_data()` function and filtering to Indiana.

```{r}
library(dplyr)

county_data <- map_data("county") %>%
  filter(region == "indiana")

head(county_data)

```

```{r}
ggplot(data = county_data) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = NA, color = "black") +
  geom_point(data = monitors, aes(x = long, y = lat)) +
  coord_fixed(1.3)
```


# ggmap

The `ggmap` package is made specifically to create `ggplot2` style maps using 
raster tiles from services like 
[Google Maps](https://developers.google.com/maps/documentation/maps-static?hl=en)
and [Stamen Maps](http://maps.stamen.com/#terrain/12/37.7706/-122.3782). __Note__
The Google Maps service now requires a key.

```{r, message=FALSE, warning=FALSE}
library(ggmap)
indpls_monitors <- monitors %>%
  filter(grepl("Indpls", name)) %>%
  rename(lon = long)

qmplot(lon, lat, data = indpls_monitors, maptype = "terrain", color = I("red"))

```

# leaflet


Leaflet is another interactive map that can be created in R using the `leaflet`
package.

```{r, warning=FALSE}
library(leaflet)
m <- leaflet()
m <- addTiles(m)
m <- addMarkers(m, lng=monitors$long, lat=monitors$lat, 
                popup=monitors$name)
m
```



