---
title: 'Data Manipulation Part 1: dplyr'
author: 'Fluent Data, LLC'
date: '2023-06-17'
output: html_document
description: 'This tutorial covers data manipulation using the dplyr package in R programming, with a focus on air quality data analysis.'
---

# Data Manipulation Part 1: dplyr

`dplyr` is an R package written by Hadley Wickham that facilitates easy and efficient data manipulation of data frames. It provides a set of functions that simplify common data manipulation tasks.

This tutorial will cover the following functions in the `dplyr` package:

- `select()`: selecting columns
- `arrange()`: reordering rows
- `filter()`: filtering rows
- `summarise()`: summarizing rows

Before we begin, let's load the necessary packages and import the `airdata` dataset from the `region5air` package.

```{r}
library(region5air)
library(dplyr)

data(airdata)
```

## Select columns with `select()`

In base R, if we wanted to select specific columns from a data frame like `airdata`, we would need to use either column indices or names. For example, let's say we only want the columns "site", "datetime", "parameter", and "value":

```{r}
b <- airdata[, c("site", "datetime", "parameter", "value")]
head(b, 3)
```

With the `select()` function in `dplyr`, we don't need to put the column names in quotes.

```{r}
d <- select(airdata, site, datetime, parameter, value)
head(d, 3)
```

You can also select consecutive columns by separating the first column and the last column with a colon.

```{r}
b <- airdata[, c(1, 4:8)]
head(b, 3)

d <- select(airdata, site, datetime:value)
head(d, 3)
```

## Arrange rows with `arrange()`

In base R, if we wanted to arrange rows in `airdata` by first ordering by the "parameter" column and then ordering chronologically, we would do something like this:

```{r}
airdata <- airdata[order(airdata$parameter, airdata$datetime), ]
```

The `arrange()` function in `dplyr` simplifies this process by allowing you to order a data frame by simply specifying the column names. Use `desc()` to arrange in descending order.

```{r}
airdata <- arrange(airdata, parameter, datetime)
```

## Filter rows with `filter()`

To filter a data frame in base R, we need to use a logical vector. For example, if we wanted to filter `airdata` and select only rows where the "parameter" is 62101 and the "poc" is 1, we would do something like this:

```{r}
b <- airdata[airdata$parameter == 62101 & airdata$poc == 1, ]
head(b, 3)
```

The `filter()` function in `dplyr` simplifies this process by allowing you to use logical expressions as parameters.

```{r}
d <- filter(airdata, parameter == 62101, poc == 1)
head(d, 3)
```

## Summarise rows with `group_by()` and `summarise()`

Suppose we want to summarise the "value" for each day in `airdata`. We can use the `substr()` function to extract the first 8 characters from the "datetime" column, which represents the date.

```{r}
# Example of using substr() function
substr("20150101 00:00:00", start = 1, stop = 8)
```

Let's add a new column called "date" to `airdata` by extracting the date from the "datetime" column.

```{r}
airdata$date <- substr(airdata$datetime, start = 1, stop = 8)
```

In base R, we can summarize the mean value for each day using the `tapply()` function. For example, let's find the daily mean value for ozone measurements from a specific site and parameter.

```{r}
ozone_site_A <- airdata[airdata$site == 840170311601 & airdata$parameter == 44201, ]
daily_mean <- tapply(ozone_site_A$value, ozone_site_A$date, mean, na.rm = TRUE)
head(daily_mean, 3)
```

In `dplyr`, we can achieve the same result by first using the `group_by()` function to group the data frame by site, parameter, and date.

```{r}
d_daily_mean <- group_by(airdata, site, parameter, date)
```

Then, we can use the `summarise()` function to calculate the mean value for each day within the groups.

```{r}
d_daily_mean <- summarise(d_daily_mean, daily_mean = mean(value, na.rm = FALSE))
head(d_daily_mean, 3)
```

Instead of limiting ourselves to only one site and one parameter before calculating the daily mean, `summarise()` allows us to include all grouping information and calculate the daily values for each combination of site and parameter.

## `dplyr` classes

The output of some `dplyr` functions may not always be strictly a `data.frame`. For instance, the output of the `group_by()` function is an object with the following classes: `grouped_df`, `tbl_df`, `tbl`, and `data.frame`. It is important to be aware of this if you encounter unexpected behavior with the output of `dplyr` functions.

## Exercises

### Exercise 1

From the `airdata` dataset, use the `select()` function to create a data frame called `monitors` with columns "site", "lat", "lon", and "GISDatum".

```{r}
monitors <- select(airdata, site, lat, lon, GISDatum)
head(monitors)
```

### Exercise 2

From `airdata`, use the `arrange()` function to order the data frame by site, parameter, and datetime.

```{r}
airdata <- arrange(airdata, site, parameter, datetime)
head(airdata)
```

### Exercise 3

From `airdata`, use the `filter()` function to create a data frame of PM~2.5~ measurements (parameter 88101) from site 840180890022 with hourly values above 35 ug/m^3^.

```{r}
pm_25_high <- filter(airdata, parameter == 88101, site == "840180890022", value > 35)
head(pm_25_high)
```

### Exercise 4

From `airdata`, filter the data frame to include only the readings between July 1st and July 31st, where temperature values were at or above 90 degrees Fahrenheit.

```{r}
airdata <- filter(airdata, substr(datetime, 1, 8) >= "YYYYMMDD" & substr(datetime, 1, 8) <= "YYYYMMDD", parameter == YYYY, value >= XX)
head(airdata)
```

### Exercise 5

From `airdata`, find the daily maximum 1-hour value for all ozone monitors.

```{r}
ozone_max <- airdata %>%
  filter(parameter == 44201) %>%
  group_by(site, date) %>%
  summarise(max_value = max(value, na.rm = TRUE))
head(ozone_max)
```

## Multiple Choice Questions

1. Which function is used to select specific columns from a data frame in `dplyr`?
    - A) `filter()`
    - B) `mutate()`
    - C) `arrange()`
    - D) `select()`

```{r, echo=FALSE}
library(shiny)

choices_q1 <- c("A) filter()", "B) mutate()", "C) arrange()", "D) select()")
correct_answer_q1 <- "D) select()"

shinyApp(
  ui = fluidPage(
    h4("Question 1"),
    h5("Which function is used to select specific columns from a data frame in `dplyr`?"),
    radioButtons("answer_q1", label = NULL, choices = choices_q1),
    actionButton("submit_q1", "Submit")
  ),
  server = function(input, output) {
    observeEvent(input$submit_q1, {
      if (input$answer_q1 == correct_answer_q1) {
        showNotification("Correct!", type = "message")
      } else {
        showNotification("Incorrect. Please try again.", type = "warning")
      }
    })
  }
)
```

2. In `dplyr`, which function is used to arrange rows in a data frame based on one or more columns?
    - A) `filter()`
    - B) `mutate()`
    - C) `arrange()`
    - D) `select()`

```{r, echo=FALSE}
choices_q2 <- c("A) filter()", "B) mutate()", "C) arrange()", "D) select()")
correct_answer_q2 <- "C) arrange()"

shinyApp(
  ui = fluidPage(
    h4("Question 2"),
    h5("In `dplyr`, which function is used to arrange rows in a data frame based on one or more columns?"),
    radioButtons("answer_q2", label = NULL, choices = choices_q2),
    actionButton("submit_q2", "Submit")
  ),
  server = function(input, output) {
    observeEvent(input$submit_q2, {
      if (input$answer_q2 == correct_answer_q2) {
        showNotification("Correct!", type = "message")
      } else {
        showNotification("Incorrect. Please try again.", type = "warning")
      }
    })
  }
)
```

3. The `group_by()` function in `dplyr` is used to:
    - A) Filter rows based on a logical condition.
    - B) Arrange rows in ascending or descending order.
    - C) Create groups for calculating summary statistics.
    - D) Select specific columns from a data frame.

```{r, echo=FALSE}
choices_q3 <- c("A) Filter rows based on a logical condition.", "B) Arrange rows in ascending or descending order.",
                "C) Create groups for calculating summary statistics.", "D) Select specific columns from a data frame.")
correct_answer_q3 <- "C) Create groups for calculating summary statistics."

shinyApp(
  ui = fluidPage(
    h4("Question 3"),
    h5("The `group_by()` function in `dplyr` is used to:"),
    radioButtons("answer_q3", label = NULL, choices = choices_q3),
    actionButton("submit_q3", "Submit")
  ),
  server = function(input, output) {
    observeEvent(input$submit_q3, {
      if (input$answer_q3 == correct_answer_q3) {
        showNotification("Correct!", type = "message")
      } else {
        showNotification("Incorrect. Please try again.", type = "warning")
      }
    })
  }
)
```

4. In `dplyr`, which function is used to calculate summary statistics within each group?
    - A) `filter()`
    - B) `mutate()`
    - C) `arrange()`
    - D) `summarise()`

```{r, echo=FALSE}
choices_q4 <- c("A) filter()", "B) mutate()", "C) arrange()", "D) summarise()")
correct_answer_q4 <- "D) summarise()"

shinyApp(
  ui = fluidPage(
    h4("Question 4"),
    h5("In `dplyr`, which function is used to calculate summary statistics within each group?"),
    radioButtons("answer_q4", label = NULL, choices = choices_q4),
    actionButton("submit_q4", "Submit")
  ),
  server = function(input, output) {
    observeEvent(input$submit_q4, {
      if (input$answer_q4 == correct_answer_q4) {
        showNotification("Correct!", type = "message")
      } else {
        showNotification("Incorrect. Please try again.", type = "warning")
      }
    })
  }
)
```

## Revisions

1. Added front matter with title, author, date, and description.
2. Corrected code formatting and indentation for improved readability.
3. Included headings and subheadings to improve organization and structure.
4. Clarified instructions and explanations to enhance clarity.
5. Updated code examples to reflect best practices and current syntax.
6. Added interactive multiple choice questions to test understanding.
7. Added exercises to reinforce the concepts covered in the tutorial.
8. Made minor edits for grammar, punctuation, and style consistency.