---
title: 'Data Manipulation Part 2: dplyr and tidyr'
output: 
  html_document: 
    keep_md: yes
---

This tutorial will cover the following main functions from [dplyr](https://dplyr.tidyverse.org/)
and [tidyr](https://tidyr.tidyverse.org/):

- `mutate()` and `transmute()`
- `pivot_longer()`
- `pivot_wider()`

It also lists some other useful miscellaneous functions.

# Add new columns with mutate() and transmute()

In the first [Data Manipulation](training-r-intermediate/blob/initial-update/1-Data-Manipulation-1/readme.md)
tutorial we learned how to use the `group_by()` and `summarise()` functions to 
aggregate data. Those functions are useful in situations where you need to 
summarise data based on other columns, such as daily summaries for each site and
pollutant in a data set. 

But in some cases we may want a new value for each record in a data frame. For example, 
in the `chicago_air` data set from the `region5air` package, we may want a column 
that is a one day lag value of the ozone column. The `mutate()` function will
take the data frame provided in the first parameter and add additional parameters
as new columns.

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(region5air)
data(chicago_air)

chicago_mutated <- mutate(chicago_air, ozone_lag1 = lag(ozone, n = 1))
head(chicago_mutated, 3)
```

Or we may want to calculate a the difference betwen the ozone value and a standard value.

```{r}
chicago_mutated <- mutate(chicago_mutated, diff_standard = ozone - 0.075)
head(chicago_mutated, 3)
```

`transmute()` works the same as `mutate()` except that it only returns the new columns.

```{r}
chicago_transmuted <- transmute(chicago_air, ozone_lag1 = lag(ozone, n = 1), 
                                diff_standard = ozone - 0.075)
head(chicago_transmuted, 3)
```

You can keep a column by including it as a parameter.

```{r}
chicago_transmuted <- transmute(chicago_air, date, ozone, lag1 = lag(ozone, n= 1))
head(chicago_transmuted, 3)
```

# Reshaping with pivot_longer()

Often we have data in a format that is not convenient for a certain purpose. For
instance, we may look at the `chicago_air` data set

```{r}
head(chicago_air)
```

and wish that we could have one column with all the values and one column with 
all of the parameter names. The current format would be called wide, because it 
has the parameters spread out as columns. Since we want to pivot those column 
names into one column and make the data frame longer, we use the `pivot_longer()`
function in the `tidyr` package.

```{r, message=FALSE, warning=FALSE}
library(tidyr)
chicago_long <- pivot_longer(chicago_air, cols = ozone:pressure, 
                             names_to = "parameter", values_to = "value")
head(chicago_long, 3)

```


One reason to reshape data in this way is to make it easier to feed into a plotting
function such as `ggplot()`.

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
ggplot(chicago_long, aes(date, value)) + geom_point() + 
  facet_grid(parameter ~., scales = "free")
```

# Reshape with pivot_wider()

We may also want to reshape data that is long into a wide format. Using the 
`chicago_long` data frame we just created, we can re-create the wide format using
`pivot_wider()`.

```{r}
chicago_wide <- pivot_wider(chicago_long, names_from = parameter,
                            values_from = value)

head(chicago_wide, 3)

```

# Miscellaneous functions

Below is a list of other useful functions from `dplyr` and `tidyr`. 

The example code utilizes the pipe `%>%`, as described in the first 
[Data Manipulation](training-r-intermediate/blob/initial-update/1-Data-Manipulation-1/readme.md)
tutorial.

## from `dplyr`

- `distinct()`: get distinct values from a data frame

```{r}

chicago_long %>%
  select(parameter) %>%
  distinct()

```

- `slice()`: select rows using a numeric vector

```{r}

chicago_air %>%
  slice(1:5)

```

- `slice_sample()`: select a certain number of rows randomly

```{r}
chicago_air %>%
  slice_sample(n = 5)

```

- `pull()`: get a column from the data frame as a vector

```{r}
ozone <- chicago_air %>%
  pull(ozone)
head(ozone)
```

## from `tidyr`

- `separate()`: separate a column into multiple columns using a delimiter

```{r, warning=FALSE, message=FALSE}
# make an example column using full site info
combined_site <- chicago_aqs %>%
  transmute(Site_ID = paste(State_Code, County_Code, Site_Number, POC, sep = "-"),
            Date_Local, Parameter_Code, AQI)

head(combined_site, 3)

# split up the Site_ID back into 4 columns
split_site <- combined_site %>%
  separate(Site_ID, c("state", "county", "site", "poc"), sep = "-")

head(split_site, 3)
  
```

- `replace_na()`: replace `NA` values in a data frame with a specific value 

```{r, warning=FALSE, message=FALSE}
# create a data frame
df <- data.frame(day = 1:5, count = c(NA, 4, 6, 2, 1), type = c("a", "b", NA, "a", "b"))

df

# provide a named list for columns and what to replace NA values with
df %>%
  replace_na(list(count = 0, type = "unknown"))
```

- `expand_grid()`: makes a data grid out of all possible combinations of input vectors

```{r, warning=FALSE, message=FALSE}
sites <- c(1:3)
dates <- seq(as.Date("2023-01-01"), as.Date("2023-01-05"), by = 1)

expand_grid(sites, dates)
```